title: "Problem Set 3 Explore One Variable"  
output: html_document  
===========================================

## Set up

Go to working directory Load the ggplot package  
```{r set things up} 
#setwd("D:/DOCUMENTS/LEARNING/DAND/DAND4 EDA - R/PS3_One_Variable")
```

--------------------------

# PS3-1, QUIZ: DIAMONDS

### Load the diamonds data set in RStudio. As a reminder, you can use these
commands to load the diamonds data set. Some data sets come with R. You can load
these data sets by typing `data(DATA_SET_NAME)`. The "diamonds" data set comes
with the *ggplot* package.

```{r load diamonds}
library(ggplot2)
data(diamonds)
```

You need to run these commands each time you launch RStudio to access the
diamonds data set. RStudio won't remember which packages and data sets you
loaded unless you change your preferences or save your workspace.

You should notice that the diamonds data set will be listed as a 'Promise' in the workspace. *Promise* is a special object in R, and you need to run a command on the data to fully load the data set.

Try running a summary: `summary(diamonds)`

You can learn more about the diamonds data set by typing `?diamonds`

Read the help file to determine the best color for the diamonds data set. When you run `diamonds$color` you will notice that the levels of color are set as "D
< E < F < ..." This is an ordering of the colors.

__How many observations are in the data set?__ `dim(diamonds)` or `?diamonds` #
preferred

__How many variables are in the data set?__ `dim(diamonds)` or `?diamonds` #
preferred

__How many ordered factors are in the data set?__ 3 `str(diamonds)`

__ What letter represents the best color for a diamond?__ I read the help file 
`?diamonds`

"color: diamond colour, from J (worst) to D (best)"

========================================================

***


# PS3-2, Price Histogram

Create a histogram of the price of all the diamonds in the diamond data set. 
TYPE YOUR CODE BELOW THE LINE  
-------------------------------

`qplot(x = price, data = diamonds)`


========================================================

***

# PS3-3, Price Histogram Summary

Describe the shape and center of the price distribution. Include summary
statistics like the mean and median.

`summary(diamonds$price)`

Shape of the distribution is skewed right with a long tail. The mean is 2401. 
The median is 3933.


========================================================

***

# PS3-4, Diamond Counts

__How many diamonds cost less that $500?__ `summary(diamonds$price < 500)` Mode 
FALSE    TRUE    NA's logical   52211    1729       0

__How many diamonds cost less that $250?__ `summary(diamonds$price < 250)` Mode 
FALSE    NA's logical   53940       0

__How many diamonds cost $15,000 or more?__ `summary(diamonds$price >= 15000)` 
Mode   FALSE    TRUE    NA's logical   52284    1656       0


========================================================

***

# PS3-5, Cheaper Diamonds

Explore the largest peak in the price histogram you created earlier.

Try limiting the x-axis, altering the bin width, and setting different breaks on
the x-axis.

Go to the discussions to share your thoughts and discover what other people
find.

You can save images by using the `ggsave()` command. `ggsave()` will save the
last plot created. For example... `qplot(x = price, data = diamonds)` 
`ggsave('priceHistogram.png')`

Submit your final code when you are ready.

TYPE YOUR CODE BELOW THE LINE -----------------------------

```{r cheaper diamonds}
# limit the x-axis, alter binwidth, set breaks 
qplot(x = price, data = diamonds, binwidth = 50, 
      color = I('black'), fill = I('#9085BA')) + 
  scale_x_continuous(limits = c(250, 1250), breaks = seq(300, 1200, 100))

ggsave('cheaperHistogram.png')
```

========================================================

***

# PS3-6, Price-by-Cut Histograms

Break up the histogram of diamond prices, by their "cut". You should have five
histograms in separate panels on your resulting plot.

TYPE YOUR CODE BELOW THE LINE -----------------------------

```{r price by cut histo}
qplot(x = price, data = diamonds, binwidth = 50) +
  scale_x_continuous(limits = c(0, 20000), breaks = seq(0, 20000, 2000)) + 
  facet_wrap(~cut, ncol = 1)
```


========================================================

***

# PS3-7, Price-by-Cut

`summary(diamonds$price)` 
Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
326    950    2401    3933    5324   18820

Which cut has the highest priced diamond?
`subset(diamonds, diamonds$price == max(diamonds$price))`
carat     cut color clarity depth table price   x    y   
z 27750  2.29 Premium     I     VS2  60.8    60 18823 8.5 8.47 5.16 
Answer: "Premium"

Which cut has the lowest priced diamond?
`subset(diamonds, diamonds$price == min(diamonds$price))` 
carat     cut color clarity depth table price    x    y   
z 1  0.23   Ideal     E     SI2  61.5    55   326 3.95 3.98 2.43 2  0.21 Premium
E     SI1  59.8    61   326 3.89 3.84 2.31 
Answer: "Premium" and "Ideal"

Which cut has the lowest median price?
`by(diamonds$price, diamonds$cut, median)`# median price, by cut
diamonds$cut: Fair [1] 3282 
diamonds$cut: Good [1] 3050.5 
diamonds$cut: VeryGood [1] 2648 
diamonds$cut: Premium [1] 3185
diamonds$cut: Ideal [1] 1810
Answer: "Ideal"

*Note: Could also have answered all 3 questions by running the following code:*
`by(diamonds$price, diamonds$cut, summary)` # summarize price, by cut

diamonds$cut: Fair Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 337    2050   
3282    4359    5206   18570 diamonds$cut: Good Min. 1st Qu.  Median    Mean 3rd
Qu.    Max. 327    1145    3050    3929    5028   18790 diamonds$cut: Very Good 
Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 336     912    2648    3982    5373
18820 diamonds$cut: Premium Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 326   
1046    3185    4584    6296   18820 diamonds$cut: Ideal Min. 1st Qu.  Median   
Mean 3rd Qu.    Max. 326     878    1810    3458    4678   18810


========================================================

***

# PS3-8, Scales and Multiple Histograms

**How to display multiple plots with different y-axis scales**

In the previous exercise, we looked at the summary statistics for diamond price
by cut.

In the output table, the median and quartiles are reasonably close to each
other. This means the distributions should be somewhat similar, but the
histograms we created don't show that. The 'Fair' and 'Good' diamonds appear to
have different distributions compared to the better cut diamonds. They seem
somewhat uniform on the left with long tails on the right.

Let's look in to this... Look up the documentation for `facet_wrap` in R Studio.
Then, scroll back up and add a parameter to `facet_wrap` so that the y-axis in
the histograms is not fixed. You want the y-axis to be different for each
histogram.

If you want a hint, check out the Instructor Notes. [ggplot2 Free scales with
facets](http://www.cookbook-r.com/Graphs/Facets_(ggplot2)/#free-scales)

ref: PS3-6 {r price by cut histo}

```{r multi scales}
ggplot(diamonds, aes(x=price)) + geom_histogram(binwidth = 50) + 
  scale_x_continuous(limits = c(0, 20000), breaks = seq(0, 20000, 2500)) +
  facet_wrap(~cut, scales='free_y')
```

Grader response: Great Job! Are your y-axis scales different on some of the
panels? It's a good idea to consider the scales when you create plots.


========================================================

***

# PS3-9, Price per Carat by Cut

Create a histogram of price per carat and facet it by cut. You can make
adjustments to the code from the previous exercise to get started.

Adjust the bin width and transform the scale of the x-axis using log10.

**Hints** Hint 1: You use the price and carat variables in the parameter for x.
What expression gives you price per carat?

Hint 2: For long tailed distributions, you can add a ggplot layer such as
`scale_x_log10()` to transform the variable.

ENTER YOUR CODE BELOW THIS LINE -------------------------------

```{r price per carat by cut}
qplot(x = price/carat, data = diamonds, binwidth = 0.1) + 
  facet_wrap(~cut, scales='free_y') + scale_x_log10()
```

**Lessons Learned:** I learned in this problem that only a binwidth under 1.0 
would produce a meaningful histogram. Sure it has somthing to do with
logarithms, and also maybe logs combined with a fraction (price/carat)-- *NAAHH*,
probably just because its logs. But note, you're only using a logarithmic scale
because the plain histogram plots *had long tails*.


========================================================

***

# PS 3-10, Price Box Plots

Investigate the price of diamonds using box plots, numerical summaries, and one
of the following categorical variables: cut, clarity, or color.

There won't be a solution video for this exercise so go to the discussion thread
for either BOXPLOTS BY CLARITY, _BOXPLOT BY COLOR_, or BOXPLOTS BY CUT to
share you thoughts and to see what other people found.

__I chose PRICE and COLOR__

**Box Plots** HINT: For boxplots, we use -- the *y* parameter for the
*continuous* data -- the *x* parameter for the *categorical* data

__*Very useful, relevant link!:*__ [Boxplots are ideally suited for visualizing
data variability](http://bxhorn.com/ggplot-boxplots/)

and maybe, 
"http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/#box-plots" 
and maybe, "http://www.statmethods.net/graphs/boxplot.html"

```{r price color boxplots} 
ggplot(diamonds, aes(x=factor(color), y=price)) + 
  geom_boxplot(aes(fill=factor(color)), outlier.size = 0.2) 
ggsave('3-10_ggboxplot_smalldot-outliers.png')
```

```{r qplot with colored boxes} 
qplot(x=factor(color), y=price, data=diamonds, geom='boxplot', fill=color) 
ggsave('3-10_qplot_box_colored.png')
```

Mark a spot at the mean (see cookbook-r.com reference, above) 
```{r boxplot with mean marked} 
qplot(x=color, y=price, data=diamonds, geom='boxplot',fill=color) + 
  stat_summary(fun.y=mean, geom="point", shape=1, size=3) 
ggsave('3-10_qplot_box_with_mean.png')
```

```{r save as above, with ggplot} 
ggplot(diamonds, aes(x=factor(color), y=price)) + 
  geom_boxplot(aes(fill=factor(color))) + 
  stat_summary(fun.y=mean, geom="point", shape=5, size=2)
```

```{r Numerical summaries}
table(diamonds$color) # count of observations by color
by(diamonds$price, diamonds$color, summary) # summarize price by color
```

**My Comments:** Min and Max prices hardly vary between colors. All of them are
more dispersed in the "high" half of prices (above median), but particularly
D-G. Less price Variance in better (D-F) colors and more variance in worse
colors (G-J.) Yet the better colors have more outliers; that might pull their
means higher; see marked-mean plots.


========================================================

***

# PS 3-11, Interquartile Range - IQR

You can use the function `IQR()` to find the interquartile range. Pass it a
subset of the diamonds data frame.

For example... `IQR(subset(diamonds, price <1000)$price)`

Remember, `subset` returns a *data frame*, so we need to use $price on the end
to access that variable.

range of middle 50% of color D diamonds?  
`by(diamonds$price, diamonds$color=='D', summary)`  
Look at the TRUE line returned.

Price range of middle 50% of color J diamonds? `by(diamonds$price,
diamonds$color=='J', summary)` Look at the TRUE line returned.

IQR of diamonds with the best color? `IQR(subset(diamonds, color=='D')$price)`

IQR of diamonds with the worst color? `IQR(subset(diamonds, color=='J')$price)` 
- - - - - - - - - - - - - - - - - - - -

alternately, IQR of all each and every color: `by(diamonds$price,
diamonds$color, IQR)`


========================================================

***

# PS 3-12, Price per Carat Box Plots by Color

Investigate the price per carat of diamonds across the different colors of
diamonds using *boxplots*.

Save images by using the ggsave() command.

```{r price-carat box by color}
qplot(x=factor(color), y = price/carat, data=diamonds, 
      geom='boxplot', fill=color, outlier.size = 0.1) 
ggsave('3-12_ppcarat_by_color_boxplots.png')
```

Comparing price per carat by color to the previous plot of price (alone) by
color, I noticed difference once you change price to price/carat: *Price per
Carat boxplots much more similar across colors than they were when we only
looked at price. *Less different medians, less different variability. *The IQRs, too, are much more similar across colors, now. *Price per Carat basically brings the "better" colors' boxplots "down" to the level of the "worse" colors' plots. *Using price per carat on y-axis removes a lot of outliers from the "better" colors.


========================================================

***

# PS 3-13 Carat Frequency Polygon

Investigate the weight of the diamonds (carat) using a frequency polygon. Use
different bin widths to see how the frequency polygon changes. What carat size
has a count greater than 2000?

```{r}
qplot(x=carat, data=diamonds, binwidth=0.1, geom='freqpoly') + 
scale_x_continuous(breaks=seq(0, 5.0, 0.1)) + 
  scale_y_continuous(breaks = seq(0, 10000, 2000))
```

For the question, we are looking for counts of exact value - try starting out
with a larger bin width and making it smaller to see how the plot changes, then focus in on the values of most interest.

**Lessons Learned:** Dealing with _**continuous** data on the x-axis_ in a
_frequency polygon_ lets you assign a __binwidth less than 1__. I'd still been
thinking in terms of _discrete_ bin counts as in a bar plot.


========================================================

***

# PS 3-15, Gapminder Data quiz

The Gapminder website contains over 500 data sets with information about the
world's population. You can find a link to the Gapminder website in the
Instructor Notes.

Your task is to download a data set of your choice and create 2-5 plots that
make use of the techniques from Lesson 3.

You might use a **simple histogram**, a **boxplot split over a categorical
variable**, or a **frequency polygon**. The choice is yours! Once you've
completed your investigation, create a post in the discussions that includes: 1. any questions you answered, your observations, and summary statistics  
2. snippets of code that created the plots  
3. links to the images of your plots  

You can save images by using the ggsave() command.

**Hints**  
The following command may be helpful for some of the Gapminder Data
sets, once it's been converted to csv format:  
`read.csv('data.csv', header = T, row.names = 1, check.names = F)`

You may want to look at additional function arguments in the help file for the
`read.table()` family for additional tools that can help you read in data.

If you want to exchange the rows and columns of your dataframe, it will also be useful to know the transpose function, t().

Copy and paste all of the code that you used for your investigation, and submit it when you are ready  
---------------------------------


========================================================

***

# PS 3-16, Exploring Friends' Birthdays

Your task is to investigate the distribution of your friends' birth months and
days.

Here some questions you could answer, and we hope you think of others.

**How many people share your birthday? Do you know them?**
**Which month contains the most number of birthdays?**
**How many birthdays are in each month?**
**Which day of the year has the most number of birthdays?**
**Do you have at least 365 friends that have birthdays on every day of the
year?**

You'll need to export the birthdays to a csv file (or use the csv file we've
provided a link to in the Instructor Notes)

Once you load the data into R Studio, you can use the `strptime()` function to
extract the birth months and birth days. We recommend looking up the 
documentation for the function and finding examples online.

- - - - - - - - -

```{r 3-16 exploring birthdays}
# Import the data file
tb <- read.csv('birthdaysExample.csv')

# tb$dates is a factor, and when I try to convert it to date it converts all
# values to NA; so, try to convert from factor to vector of characters
tb$dates <- as.character(tb$dates)

# Convert to date format
tb$monthDay <- as.Date(tb$dates, format = "%m/%d")

# Month column
tb$month <- as.numeric(format(tb$monthDay, format = '%m'))

# Day column
tb$day <- as.numeric(format(tb$monthDay, format = '%d'))

# Count my birthday
myBday <- ifelse(tb$dates == '5/23/14', 1, 0)
sum(myBday)
```

Month with most birthdays?
```{r}
qplot(x = month, data = tb, binwidth = 1) + 
  scale_x_continuous(breaks = seq(1, 12, 1))
```

Answer: March


How many birthdays are in each month?  
by(tb$monthDay, tb$month, FUNCTION???)  
```{r}
ggplot(tb, aes(x=month), labels = ..count..) + 
  geom_bar() + 
  stat_bin(geom="text", aes(label=..count.., vjust=-1))
ggsave('birthdays per month.png')
```

Which day of the year has the most number of birthdays?  
(I found two methods online)  
```{r most bdays 1}
datefreq <- data.frame(table(tb$monthDay))
datefreq[which.max(datefreq$Freq), ]
```
         Var1 Freq  
37 2016-02-06    8  

Answer: February 6th.


